---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

options(scipen = 999)
```

# opmitools

<!-- badges: start -->

<!-- badges: end -->

`opmitools` serves 3 primary purposes:

1.  Functions for cleaning and merging GTFS files (building on [`tidytransit`](https://github.com/r-transit/tidytransit))
2.  Presenting data consistent with OPMI's Office Theme (building on [`ggplot2`](https://ggplot2.tidyverse.org/))
    1.  This part of the package draws on "[Creating corporate colour palettes for ggplot2](https://drsimonj.svbtle.com/creating-corporate-colour-palettes-for-ggplot2)"
3.  Access to common reference tables

## Installation

You can install the development version of opmitools from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("kmeakinmbta/opmitools")
```

And load/attach opmitools as you would any other package with:

```{r}
library(opmitools)
```

## Usage

### GTFS Files

Read the current MBTA GTFS from [mbta.com](https://www.mbta.com/) with:

```{r, message=FALSE}
mbta_gtfs <- read_mbta_gtfs()
```

`read_mbta_gtfs()` is a wrapper for `tidytransit::read_gtfs()`. It stores the individual tables in a list and enforces column types as specified in the GTFS standard.

For some downstream trip planners and other applications, it is useful/necessary to remove all extraneous stops, shapes, routes, and `service_id`s (i.e., those that are not used in the either `trips.txt` or `stop_times.txt`). You can check for extraneous components with `gtfs_removal_check()`:

```{r}
gtfs_removal_check(mbta_gtfs)
```

And remove those components with `gtfs_remove_all()`:

```{r}
route_count_pre <- length(mbta_gtfs$routes$route_id)
mbta_gtfs <- gtfs_remove_all(mbta_gtfs)
route_count_post <- length(mbta_gtfs$routes$route_id)
paste0("Route count prior to removal: ", route_count_pre)
paste0("Route count after removal: ", route_count_post)
```

Other use cases might include getting some useful information out of the `calendar.txt` file:

```{r}
mbta_calendar_info <- gtfs_calendar_info(mbta_gtfs)
head(mbta_calendar_info)
```

Note that `opmitools` is not yet capable of handling GTFS-Pathways -- it is only set up to handle the required GTFS files used by the MBTA: agency, calendar, routes, shapes, stop_times, stops, and trips. While some functions will work without `tidytransit`, it is recommended to install and attach `tidytransit` in conjunction with `opmitools`.

### Plotting

`opmitools` contains several convenience functions for making "OPMI-themed" plots. As an example, we can plot information from `rt_ridership`, which is included in the package.

The following code creates a simple ggplot showing fall average weekday ridership on rapid transit.

```{r, message=FALSE}
library(tidyverse) # load & attach tidyverse for data wrangling with dplyr and plotting with ggplot2

(ridership_plot <-
    rt_ridership %>%  
    filter(day_type_name == "weekday") %>% 
    ggplot(aes(x = rating, y = tot_ons, fill = route_id)) +
    geom_col(position = position_dodge()))
```

Then, we can add the OPMI theme with `theme_opmi()`:

```{r}
(ridership_plot <-
   ridership_plot +
   labs(x = "Rating", y = "Average\nWeekday\nBoardings") +
   theme_opmi())
```

And use colors that make sense with `scale_fill_opmi()`:

```{r}
ridership_plot +
  scale_fill_opmi(palette = "rt", guide = "none")
```


### Included Data

`opmitools` contains several reference tables that may be useful, including a table describing the MBTA service area:

```{r}
head(mbta_service_area)
```

And frequently used demographic information by MA block group:

```{r}
head(mabgs_19)
```

